Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-06-26 unimodular-covers.tex
added proof of offending lemma for index two

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

@@

\begin{figure}[htb]
  \centering
\begin{tikzpicture}[mynode/.style={circle, inner sep=1.5pt, fill=red}, scale =
  .8]
%coordinate axes
   \tkzInit[xmax=3.5,ymax=3.5,xmin=-3.5,ymin=-3.5]
%        \tkzLabelX[orig=false,label options={font=\tiny}]
%        \tkzLabelY[orig=false,label options={font=\tiny}]
        \tkzGrid
%       \tkzDrawX
%       \tkzDrawY

@@

  \draw[black] (A) -- (C) -- (D) -- (F) -- (A);
  \draw[black, thick] (A) -- (D);
\end{tikzpicture}
%
\begin{tikzpicture}[mynode/.style={circle, inner sep=1.5pt, fill=red}, scale =
  .8]
%coordinate axes
   \tkzInit[xmax=3.5,ymax=3.5,xmin=-3.5,ymin=-3.5]
%        \tkzLabelX[orig=false,label options={font=\tiny}]
%        \tkzLabelY[orig=false,label options={font=\tiny}]
        \tkzGrid
%        \tkzDrawX
%        \tkzDrawY

@@

  \draw[black] (A) -- (B) -- (C) -- (D) -- (E) -- (F) -- (A);
  \draw[black, thick] (A) -- (D);
\end{tikzpicture}

@@

\caption{Partial counter-example to \Cref{lemma:cayley}. It is not a counter-example because (a) the edges $p_1p_2$ and $q_1q_2$ are not primitive and (b) the origin is a lattice point in the interior of both bands (but it is the only one). 
%
With respect to the lattice of index two it is not a counter-example either, because two vertices of $Q$ are not in that lattice.
}
\label{fig:width3}
\end{figure}

\begin{lemma}
\label{lemma:cayley}
Let $P$ and $Q$ be two lattice polygons with $Q$ nef with respect to $P$. Let $p_1p_2 \subset P$ and $q_1q_2\subset Q$ be two primitive and non-parallel segments, and assume that the prod-simplex (parallelogram)  $p_1p_2 + q_1q_2$ is not unimodular. Then,
at least one of the strips
\paco{add picture for these strips, and perhaps for the proof}
\[
(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle) \cap P
\qquad \text{and} \qquad
(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle) \cap Q
\]
has a lattice point in its interior.
\end{lemma}
\giulia{what I want is: a lattice point $p \in P$ such that either $p+q_1$ or $p+q_2$ lies in the interior of $\{p_1, p_2\} + \{q_1, q_2\}$ (or symmetric for $Q$)}

\begin{proof}
Suppose by contraddiction that there is no lattice point as described in the lemma. Then there cannot be a vertex of $P$ in the corresponding strip, since $P$ is a lattice polygon. The same holds for $Q$ and its corresponding strip. Thus $P$ has two edges which each have one vertex on each side of the strip $(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle)$; we will call these edges $e_1$ and $e_2$. The same holds for $Q$ and the strip $(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle)$, and we call the edges be $f_1, f_2$.

@@

>>>>

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{claim}[theorem]{Claim}

@@

\begin{figure}[htb]
  \centering
\begin{tikzpicture}[mynode/.style={circle, inner sep=1.5pt, fill=red}, scale =
  .7]
%coordinate axes
   \tkzInit[xmax=3.5,ymax=3.5,xmin=-3.5,ymin=-3.5]
%        \tkzLabelX[orig=false,label options={font=\tiny}]
%        \tkzLabelY[orig=false,label options={font=\tiny}]
        \tkzGrid
%       \tkzDrawX
%       \tkzDrawY

@@

  \draw[black] (A) -- (C) -- (D) -- (F) -- (A);
  \draw[black, thick] (A) -- (D);
\end{tikzpicture}
%
\begin{tikzpicture}[mynode/.style={circle, inner sep=1.5pt, fill=red}, scale =
  .7]
%coordinate axes
   \tkzInit[xmax=3.5,ymax=3.5,xmin=-3.5,ymin=-3.5]
%        \tkzLabelX[orig=false,label options={font=\tiny}]
%        \tkzLabelY[orig=false,label options={font=\tiny}]
        \tkzGrid
%        \tkzDrawX
%        \tkzDrawY

@@

  \node[mynode] (p1) at (-1,0) {}; 
  \node[mynode] (p2) at (1,0) {};

  \draw[black] (A) -- (B) -- (C) -- (D) -- (E) -- (F) -- (A);
  \draw[black, thick] (p1) -- (p2);
\end{tikzpicture}

@@

\caption{Partial counter-example to \Cref{lemma:cayley}. With respect to the standard lattice it is not a counter-example because (a) the edges $p_1p_2$ and $q_1q_2$ are not primitive and (b) the origin is a lattice point in the interior of both bands (but it is the only one). 
%
With respect to the lattice of index two it is not a counter-example either, because two vertices of $Q$ are not in that lattice.
}
\label{fig:width3}
\end{figure}

\begin{lemma}
\label{lemma:cayley}
Let $P$ and $Q$ be two lattice polygons with $Q$ nef with respect to $P$. Let $p_1p_2 \subset P$ and $q_1q_2\subset Q$ be two primitive and non-parallel segments, and assume that the prod-simplex (parallelogram)  $p_1p_2 + q_1q_2$ is not unimodular. Then,
at least one of the strips
\paco{add picture for these strips, and perhaps for the proof}
\[
(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle) \cap P
\qquad \text{or} \qquad
(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle) \cap Q
\]
has a lattice point in its interior.
\end{lemma}
\giulia{what I want is: a lattice point $p \in P$ such that either $p+q_1$ or $p+q_2$ lies in the interior of $\{p_1, p_2\} + \{q_1, q_2\}$ (or symmetric for $Q$)}

\begin{proof}
Suppose by contradiction that there is no lattice point as described in the lemma. Then there cannot be a vertex of $P$ in the corresponding strip, since $P$ is a lattice polygon. The same holds for $Q$ and its corresponding strip. Thus $P$ has two edges which each have one vertex on each side of the strip $(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle)$; we will call these edges $e_1$ and $e_2$. The same holds for $Q$ and the strip $(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle)$, and we call the edges be $f_1, f_2$.

@@

\appendix
\section{Attempt at the proof of \Cref{lemma:cayley}}

At some point I'll assume volume two for the parallelogram $p_1p_2 + q_1q_2$, but for the time being we don't.

Let us fix coordinates in which $p_1p_2$ and $q_1q_2$ are a vertical and horizontal segment of length one intersecting at the origin. That is, there are numbers $\alpha, \beta \in (0,1)$ such that
\[
p_1=(0,-\alpha), \quad
p_2=(0,1-\alpha), \quad
q_1=(-\beta,0), \quad
q_2=(1-\beta,0).
\] 
The ambient lattice is a certain lattice containing $p_1,p_2,q_1,q_2$ (the case of index two is when $\alpha=\beta=1/2$ and the lattice is spanned by these four points).

Suppose by contradiction that there is no lattice point as described in the lemma. Then there cannot be a vertex of $P$ in the corresponding strip, since $P$ is a lattice polygon. The same holds for $Q$ and its corresponding strip. Thus $P$ has two primitive edges which each have one vertex on each side of the strip $(p_1p_2 + \langle \overrightarrow{q_1q_2}\rangle)$; we will call these edges $\ell$ and $r$ (with $\ell$ to the left and $r$ to the right). The same holds for $Q$ and the strip $(q_1q_2 + \langle \overrightarrow{p_1p_2}\rangle)$, and we call the edges  $b, t$ (with $b$ below and $t$ above). To make things more explicit, we assume the following coordinates for the edge-vectors:
\[
\vec l = (l_1,l_2), \quad
\vec r= (r_1,r_2), \quad
\vec b= (b_1,b_2), \quad
\vec t= (t_1,t_2).
\]
Without loss of generality we assume $l_2,r_2,b_1,t_1 >0$. (They are non-zero and we can arbitrarily choose a sign for each vector).

Then:

\begin{claim}
$l_2, r_2, b_1, t_1 >1$.
\end{claim}

\begin{proof}
$l_2, r_2, b_1, t_1 \ge 1$ follows from the fact that each edge crosses the corresponding strip. If one of the inequalities is not strict then one of the strips contains half of the square $p_1p_2+q_1q_2$, hence it contains a lattice point in its interior.  
\end{proof}

\begin{claim}
$b_2$ and $t_2$ are non-zero and have the same sign. W.l.o.g.~we assume they are positive.
\end{claim}

\begin{proof}
If, say, $b_2 \le 0 \le t_2$. Then the intersections of $t$ and $b$ with the vertical line $\{x=-\beta\}$ through $q_1$ are less than distance one apart (because $t$ must go below every lattice point in the interior of $p_1p_2 + q_1q_2 - p_1$ and $b$ above every latttice point in the interior of $p_1p_2 + q_1q_2 - p_2$). This implies the vertical distance between any two points in $Q \cap \{x\le  -\beta\}$ is less than one, so it is impossible to fit a translated copy of $l$ in it.
\end{proof}

\begin{claim}
Assume w.l.o.g.~that $b$ has greater slope than $t$. Then, 
\begin{enumerate}
\item The intersection of $Q$ with any vertical line to the right of $q_2$ has length strictly smaller than one.
\item $r_1>0$.
\end{enumerate}
\end{claim}

\begin{proof}
The assumption implies that the intersections of $t$ and $b$ with the vertical line $\{x=1-\beta\}$ through $q_2$ are less than distance one apart (consider a vertical lattice line intersecting $q_1q_2$ and two consecutive lattice points in it, one above and one below the horizontal axis). This in turn implies part (a). For (b), if $r_1$ was $\le 0$ it would be impossible to fit a translated copy of $r$ in the right side of $Q$, because of (a).
\end{proof}

As a conclusion of the last two claims we have that $b$, $t$ and $r$ have positive slope.

Starting here we assume index 2.

\begin{claim}
The slopes of $t$ and $b$ are strictly smaller than one and that of $r$ strictly greater than one.
\end{claim}

\begin{proof}
Otherwise the points $(0,1/2)$, $(0,-1/2)$, $(1/2,0)$, respectively, would be in the corresponding band.
\end{proof}

We are now ready to show a contradiction. Let $u$ and $v$ be the end-points of the translated copy of $r$ in $Q$, with $\vec r=v-u$. Let $d$ be the line of slope $1$ through $u$, which is a lattice line. Let $d'$ be the next parallel lattice line above $d$. By Claim A.3.1, the intersection of $d'$ with the vertical line through $u$ is above $Q$. Since the slope of $t$ is less than one, the same happens for every point of $d'$ to the right of $u$. In particular, $v$ must be below $d'$ and hence it must be on or below $d$. This contradicts the fact that the slope of $r$ is strictly greater than one. QED
